---
title: "Gene Expression Analysis Report"
subtitle: "Tidyverse-based exploratory analysis of simulated RNA-seq data"
author: "tidyverse_biodata project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center"
)
```

---

## 1 · Introduction

This report presents an exploratory analysis of a simulated **gene expression
dataset** that mimics bulk RNA-seq counts normalised to log₂ scale. The
dataset contains **50 genes** (47 protein-coding, 3 lncRNA) measured across
**6 samples**: three biological replicates of a *Control* condition and three
of a *Treatment* condition.

All data manipulation is performed with [**tidyverse**](https://www.tidyverse.org/)
tools and visualisations are created with
[**ggplot2**](https://ggplot2.tidyverse.org/).

---

## 2 · Setup

```{r libraries}
library(tidyverse)
library(ggrepel)
```

---

## 3 · Data Loading

```{r load-data}
gene_expr <- read_csv(here::here("data", "gene_expression.csv"),
                      show_col_types = FALSE)

glimpse(gene_expr)
```

The dataset has `r nrow(gene_expr)` genes and `r ncol(gene_expr)` columns.
The expression columns represent log₂-normalised values; higher values
indicate greater transcript abundance.

```{r missing-values}
# Confirm no missing values
colSums(is.na(gene_expr))
```

### Gene biotype breakdown

```{r biotype-counts}
gene_expr |>
  count(gene_biotype, name = "n_genes") |>
  knitr::kable(caption = "Gene biotype distribution")
```

---

## 4 · Data Wrangling

We reshape the wide-format data to **tidy (long) format** — one observation
per row — which makes downstream summarisation and visualisation straightforward.

```{r pivot-long}
gene_expr_long <- gene_expr |>
  pivot_longer(
    cols      = ctrl_rep1:treat_rep3,
    names_to  = "sample",
    values_to = "expression"
  ) |>
  mutate(
    condition = if_else(str_starts(sample, "ctrl"), "Control", "Treatment"),
    replicate = str_extract(sample, "\\d+$"),
    .after    = sample
  )

head(gene_expr_long)
```

### Log₂ fold change

We compute the **mean expression per gene per condition** and then derive
the log₂ fold change (Treatment − Control, since values are already on the
log₂ scale).

```{r fold-change}
gene_means <- gene_expr_long |>
  group_by(gene_id, gene_name, gene_biotype, condition) |>
  summarise(mean_expr = mean(expression), .groups = "drop")

gene_fc <- gene_means |>
  pivot_wider(names_from = condition, values_from = mean_expr) |>
  mutate(
    log2fc     = Treatment - Control,
    abs_log2fc = abs(log2fc),
    direction  = case_when(
      log2fc >  1 ~ "Up",
      log2fc < -1 ~ "Down",
      TRUE        ~ "Stable"
    )
  )

head(gene_fc)
```

---

## 5 · Summary Statistics

### 5.1 Per-sample statistics

```{r sample-stats}
sample_summary <- gene_expr_long |>
  group_by(sample, condition) |>
  summarise(
    n_genes = n(),
    mean    = round(mean(expression), 3),
    median  = round(median(expression), 3),
    sd      = round(sd(expression), 3),
    .groups = "drop"
  ) |>
  arrange(condition, sample)

knitr::kable(sample_summary,
             caption = "Per-sample expression summary statistics")
```

### 5.2 Most variable genes

The **coefficient of variation** (CV = σ / μ) across all six samples
identifies the genes with the most dynamic expression changes.

```{r variable-genes}
gene_summary <- gene_expr_long |>
  group_by(gene_id, gene_name, gene_biotype) |>
  summarise(
    mean_all = round(mean(expression), 3),
    sd_all   = round(sd(expression), 3),
    cv       = round(sd_all / mean_all, 3),
    .groups  = "drop"
  ) |>
  slice_max(cv, n = 10)

knitr::kable(gene_summary,
             caption = "Top 10 most variable genes (by coefficient of variation)")
```

### 5.3 DEG summary

Using |log₂FC| > 1 as a threshold:

```{r deg-summary}
gene_fc |>
  count(direction, name = "n_genes") |>
  knitr::kable(caption = "DEG direction counts (|log2FC| > 1 threshold)")
```

---

## 6 · Visualisations

### 6.1 Expression distribution per sample

A boxplot provides a quick quality-control view — we expect similar
distributions across replicates within the same condition.

```{r boxplot, fig.height=5}
ggplot(gene_expr_long,
       aes(x = sample, y = expression, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  scale_fill_manual(values = c(Control = "#4393C3", Treatment = "#D6604D")) +
  labs(
    title    = "Gene Expression Distribution per Sample",
    subtitle = "Log2-normalised expression across all 50 genes",
    x        = "Sample", y = "Expression (log2)", fill = "Condition"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 6.2 Volcano plot

The volcano plot shows the relationship between fold change magnitude and
direction, highlighting genes with large shifts (|log₂FC| ≥ 3).

```{r volcano, fig.height=5}
volcano_data <- gene_fc |>
  mutate(label = if_else(abs_log2fc >= 3, gene_name, NA_character_))

ggplot(volcano_data,
       aes(x = log2fc, y = abs_log2fc, colour = direction, label = label)) +
  geom_point(alpha = 0.8, size = 2.5) +
  geom_text_repel(size = 3, max.overlaps = 15, na.rm = TRUE) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c(Up = "#D6604D", Down = "#4393C3", Stable = "grey60")
  ) +
  labs(
    title  = "Volcano Plot: Treatment vs Control",
    x      = "Log2 Fold Change",
    y      = "|Log2 Fold Change|",
    colour = "Direction"
  ) +
  theme_bw(base_size = 12)
```

### 6.3 Heatmap of the 20 most variable genes

```{r heatmap, fig.height=7}
top20_ids <- gene_expr_long |>
  group_by(gene_id) |>
  summarise(cv = sd(expression) / mean(expression), .groups = "drop") |>
  slice_max(cv, n = 20) |>
  pull(gene_id)

heatmap_data <- gene_expr |>
  filter(gene_id %in% top20_ids) |>
  pivot_longer(ctrl_rep1:treat_rep3,
               names_to = "sample", values_to = "expression")

ggplot(heatmap_data,
       aes(x = sample,
           y = fct_reorder(gene_name, expression),
           fill = expression)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low      = "#4393C3",
    mid      = "white",
    high     = "#D6604D",
    midpoint = median(heatmap_data$expression)
  ) +
  labs(
    title = "Heatmap: Top 20 Most Variable Genes",
    x     = "Sample", y = "Gene", fill = "Expression\n(log2)"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 6.4 PCA of sample expression profiles

```{r pca, fig.height=5}
pca_matrix <- gene_expr |>
  select(gene_id, ctrl_rep1:treat_rep3) |>
  column_to_rownames("gene_id") |>
  t()

pca_result <- prcomp(pca_matrix, center = TRUE, scale. = TRUE)

var_exp <- summary(pca_result)$importance["Proportion of Variance", ] * 100

pca_df <- as_tibble(pca_result$x[, 1:2], rownames = "sample") |>
  mutate(condition = if_else(str_starts(sample, "ctrl"), "Control", "Treatment"))

ggplot(pca_df, aes(x = PC1, y = PC2, colour = condition, label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text_repel(size = 3.5) +
  scale_colour_manual(values = c(Control = "#4393C3", Treatment = "#D6604D")) +
  labs(
    title  = "PCA of Gene Expression Profiles",
    x      = sprintf("PC1 (%.1f%% variance)", var_exp[1]),
    y      = sprintf("PC2 (%.1f%% variance)", var_exp[2]),
    colour = "Condition"
  ) +
  theme_bw(base_size = 12)
```

### 6.5 DEG counts by gene biotype

```{r bar-biotype, fig.height=4}
gene_fc |>
  filter(direction != "Stable") |>
  count(direction, gene_biotype) |>
  ggplot(aes(x = gene_biotype, y = n, fill = direction)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(Up = "#D6604D", Down = "#4393C3")) +
  labs(
    title = "Differentially Expressed Genes by Biotype",
    x     = "Gene Biotype", y = "Number of DEGs", fill = "Direction"
  ) +
  theme_bw(base_size = 12)
```

---

## 7 · Key Findings

```{r findings-table}
top_up <- gene_fc |>
  filter(direction == "Up") |>
  slice_max(log2fc, n = 5) |>
  select(gene_name, gene_biotype, Control, Treatment, log2fc) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

top_down <- gene_fc |>
  filter(direction == "Down") |>
  slice_min(log2fc, n = 5) |>
  select(gene_name, gene_biotype, Control, Treatment, log2fc) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

knitr::kable(top_up,   caption = "Top 5 upregulated genes in Treatment")
knitr::kable(top_down, caption = "Top 5 downregulated genes in Treatment")
```

**Summary:**

- **`r sum(gene_fc$direction == "Up")`** genes are upregulated in the
  Treatment condition (log₂FC > 1).
- **`r sum(gene_fc$direction == "Down")`** genes are downregulated
  (log₂FC < −1).
- **`r sum(gene_fc$direction == "Stable")`** genes show no appreciable
  change.
- The PCA clearly separates Control and Treatment samples along **PC1**,
  confirming a strong condition effect.
- Housekeeping genes (*ACTB*, *GAPDH*) show near-zero fold changes as
  expected.

---

## 8 · Session Info

```{r session-info}
sessionInfo()
```
